---
title: "Code & data representation"
output: pdf_document
---

#### Importing libraries:
```{r}
library(tidyverse)
library(lme4)
library(sjstats)
library(ggplot2)
```

#### Reading the source data:
```{r}
Sys.setlocale(category = "LC_ALL", locale = "English")
data <- read.csv("final.csv", encoding = 'UTF-8')
Sys.setlocale(category = "LC_ALL", locale = "Japanese")
```

#### Section 1: Conjugation
##### Testing for dependence:
```{r}
fisher.test(table(data$conj, data$form))
```
##### Running the regression:
```{r}
data %>%
  lm(as.numeric(form) ~ conj, data = .) %>%
  summary()
```

#### Section 2: Context postition
##### Testing for dependence:
```{r}
# without ambiguous contexts
data.unamb_cont <- data[data$context_pos != "unclear", ]
fisher.test(table(data.unamb_cont$context_pos, data.unamb_cont$form))

# with ambiguous contexts
fisher.test(table(data$context_pos, data$form))
```

##### Running the regression:
```{r}
# without ambiguous contexts
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos, data = .) %>%
  summary()

# with ambiguous contexts
data %>%
  lm(as.numeric(form) ~ context_pos, data = .) %>%
  summary()
```

#### Section 2.5: Conjugation + context postition
##### Running the regression:
```{r}
# without ambiguous contexts
data.unamb_cont %>%
  glmer(form ~ context_pos + (1|conj),
        family = binomial(link="logit"), data = ., control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun=10000))) -> gl2.5
summary(gl2.5)
icc(gl2.5)

# with ambiguous contexts
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + conj, data = .) %>%
  summary()
```

#### Section 3: Sex
##### Testing for dependence:
```{r}
fisher.test(table(data$sex, data$form))
```

##### Running the regression:
```{r}
data %>%
  lm(as.numeric(form) ~ sex, data = .) %>%
  summary()
```

#### Section 3.5: Sex + context position + conjugation
##### Running the regression:
```{r}
# without ambiguous contexts
data.unamb_cont %>%
  lm(as.numeric(form) ~ sex + context_pos + conj, data = .) %>%
  summary()

# with ambiguous contexts
data %>%
  lm(as.numeric(form) ~ sex + context_pos + conj, data = .) %>%
  summary()

# without ambiguous contexts and with conjugation as a random effect
data.unamb_cont %>%
  glmer(form ~ sex + context_pos + (1|conj),
        family = binomial(link="logit"), data = ., control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun=10000))) -> gl3.5

summary(gl3.5)
icc(gl3.5)
```

#### Section 4: Length
##### Testing for dependence:
```{r}
fisher.test(table(data$length, data$form))
```

#### Section 4.5: Length + context position + sex + conjugation
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + length + conj, data = .) %>%
  summary()
```


#### Section 5: Part of speech
##### Testing for dependence:
```{r}
fisher.test(table(data$part_of_speech, data$form))
```

##### Running the regression:
```{r}
data %>%
  lm(as.numeric(form) ~ part_of_speech, data = .) %>%
  summary()
```


#### Section 5.5: Part of speech + context position + sex + conjugation
##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + conj, data = .) %>%
  summary()
```

#### Section 6: Positivity & negativity
##### Testing for dependence:
```{r}
fisher.test(table(data$pos_neg, data$form))
```

##### Running the regression:
```{r}
data %>%
  lm(as.numeric(form) ~ pos_neg, data = .) %>%
  summary()
```


#### Section 6.5: Positivity & negativity + part of speech + context position + sex + conjugation
##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + conj, data = .) %>%
  summary()
```

#### Section 7: Past tense
##### Testing for dependence:
```{r}
fisher.test(table(data$past_tense, data$form))
```

##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ past_tense, data = .) %>%
  summary()
```


#### Section 7.5: Past tense + context position + sex + part of speech + conjugation + positivity & negativity
##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + conj + past_tense, data = .) %>%
  summary()

data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + past_tense, data = .) %>%
  summary()
```

#### Section 8: Provisional & conditional form
##### Testing for dependence:
```{r}
fisher.test(table(data$prov_cond, data$form))
```

##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ prov_cond, data = .) %>%
  summary()
```

#### Section 8.5: Provisional & conditional form + context position + sex + part of speech + past tense + positivity & negativity
##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + past_tense + prov_cond, data = .) %>%
  summary()
```

#### Section 9: Question
##### Testing for dependence:
```{r}
fisher.test(table(data$quest, data$form))
```

#### Section 9.5: Question + provisional & conditional form + context position + sex + part of speech + past tense + positivity & negativity
##### Running the regression:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + past_tense + prov_cond + quest, data = .) %>%
  summary()
```
## Final model:
```{r}
data.unamb_cont %>%
  lm(as.numeric(form) ~ context_pos + sex + part_of_speech + pos_neg + past_tense + prov_cond, data = .) %>%
  summary()
```

### Visualizing data
```{r}
data %>%
  count(length) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(length))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  labs(title = "Distribution of length (in syllables)")
data %>%
  count(context_pos) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(context_pos))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  labs(title = "Distribution of conjugations")
data %>%
  count(part_of_speech) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(part_of_speech))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("adjective", "adverb", "participle", "verb")) +
  labs(title = "Distribution of parts of speech")
data %>%
  count(pos_neg) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(pos_neg))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("negative", "positive")) +
  labs(title = "Distribution of positivity & negativity")
data %>%
  count(past_tense) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(past_tense))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("non-past", "past")) +
  labs(title = "Distribution of past tense")
data %>%
  count(conj) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(conj))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("1st", "2nd")) +
  labs(title = "Distribution of conjugations")
data %>%
  count(sex) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(sex))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("female", "male")) +
  labs(title = "Distribution of speakers' sex")
data %>%
  count(form) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(form))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("substantivation", "verbal form")) +
  labs(title = "Distribution of strategies")
data %>%
  count(prov_cond) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(prov_cond))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "vertical", legend.position="bottom") +
  scale_fill_discrete(labels = c("not a prov/cond form", "form expressed through separate word", "form expressed through a marker")) +
  labs(title = "Distribution of provisional and conditional forms")
data %>%
  count(quest) %>%
  ggplot(., aes(x="", y=n, fill=as.factor(quest))) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme(axis.title=element_blank(), legend.title = element_blank(), legend.direction = "horizontal", legend.position="bottom") +
  scale_fill_discrete(labels = c("not a question", "a question")) +
  labs(title = "Distribution of questions")
```